Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-06-01 paper/manuscript.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% What simulation protocol should I choose   %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{\tocorange{What simulation protocol should I choose?}}
\tocsectioncomment{Steering group: Possibly expand to include other methods group agrees on.}
\tocsectioncomment{@Stefan Boresch to restructure as per his suggestion to remove single/dual topology section and begin with a more fundamental discussion which might more naturally include all options e.g. ATM. @Stefan Boresch to mention best practices for dummy atoms? This might benefit from following @Emilio Gallicchio's changes to make the theory section more general.}
\tocsectioncomment{@Sheenam Khuttan / @Emilio Gallicchio to contribute section on ATM.}
\tocsectioncomment{@Vytas Gapsys / @Volunteer to add best practices for non-equilibrium calculations.}
\label{sec:simulation_protocol_choice}
Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (see Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative''\footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state and in that sense are relative, while neither computes an absolute free energy.} (see Fig.~\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
To use binding as a concrete example, in absolute binding free energy calculations, we compute the binding free energy of a ligand to an individual receptor relative to a standard reference concentration. In contrast, in relative binding free energy calculations, we compare the binding free energy of two related ligands to determine the potency difference.

@@

\paragraph{Topologies} In regular biomolecular simulations one studies physical systems or, in other words, all entities in the system, e.g., protein, ligand and waters, represent real molecules. For the following, it is helpful to classify such simulation systems as having a {\em chemical} topology. In relative free energy calculations, however, one has to transform, e.g., ligand A into ligand B etc. (see {\bf XXX}). The required alchemical transformations can be realized in several ways, each with distinct advantages and disadvantages. In the first relative free energy calculations, these transformations were achieved with the help of chimeric entities, capable of representing both ligand A or ligand B. The important point here is that there is a single set of coordinates for this chimeric ligand, but its ``topology'' is flexible and can, for the same set of coordinates, represent either end state. We will consider this class of approaches as using an {\em alchemical} topology with a single set of coordinates. The construction of these chimeric intermediates quickly gets complicated and is difficult to automate; therefore, approaches have been developed in which both end states (e.g., ligand A and B) are present simultaneously. Thus, we have dual coordinates, but chemical topologies for each of the entities. In the following, we outline these approaches in more detail, together with additional considerations how the various approaches differ in practical use. While we will point out the respective advantages and disadavantages, we stress that the user's option in practice often are limited by the choice of simulation software.

@@

In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by first changing one of the hydrogens into a carbon. The site of this transformation will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'', which retain their bonded interactions but do not interact with the rest of the system.  Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final state; see below for further details. Thus, in a single topology calculation, atoms may change their type, ensuring minimal dummy atoms are created. This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}.

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works. Note that the two groups that are switched on or off never see each other.

\begin{figure}
    \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
    \caption{\textbf{Strategies for alchemical calculations: single coordinate single and dual topology vs.\ dual coordinate dual topology.} \textbf{Left}: A single coordinate single topology converts from one type of atom to an other. Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms, using a soft-core interaction to improve overlap between the dummy atoms and the "real" atoms.  \textbf{Middle}: The single coordinate dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses soft-core potentials for this. The 'mixed' intermediate atoms (M) are used in both single coordinate approaches (single and dual topology). Only the way the transformation occurs and the end states differ. Following the arrow along the left and right illustrate the differences. \textbf{Right}: In the dual coordinate dual topology approach both end states are present as separate molecules simultaneously.  Initially, benzyl alcohol is not interacting with the environment, at intermediate states the environment sees a mixture of the interactions of the two solutes, and, finally, benzene is not interacting with the environment. The two solutes are kept on top of each other by suitable restraints, indicated by the black ``spring'', and  the two molecules never see each other.  Figure adapted from the one used in V1 of this manuscript; adapted and extended by Sara Tkaczyk.
      % \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}
    }
    \label{fig:fig_topology}
\end{figure}

\sbnote{This probably needs to be rewritten\ldots} As far as we are aware, the distinction between (single coordinate) single and dual topology approaches has been made primarily for two main reasons. First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches. Additionally, historically, some simulation packages implemented only one approach and not the other, meaning that the distinction was functional. For an in-depth analyis of differences resulting from the use of a  single vs.\ a dual topology approach (single coordinate in both cases) see Ref.~\cite{Boresch_2002}.

@@

The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts \sbnote{See below; where do we place the dummy atom discussion?}. Some of these difficulties might be avoided by having the initial and final state present as separate molecules with separate coordinate sets. To the best of our knowledge, such an approach was first described by Axelsen and Li \cite{Axelsen_1998}. Both initial and final state are present simultaneously; the two copies never see each other and interactions with the environment are turned on/off as a function of $\lambda$. Harmonic restraints between corresponding atoms in the initial and final state copy prevent the two molecules (sets of atoms from drifting apart). %Unfortunately, Axelsen and Li referred to their ansatz also as dual-topology, though in practice it is quite different from Pearlman’s meaning of the term.
A recent implementation of this idea is Restraintmaker \cite{Ries_2022}, a tool to automatically assign the distance restraints needed to keep the two molecules / regions (one for each of the physical endstates) on top of each other. Somewhat unfortunately, Axelsen and Li, as well as Ries et al refer to their approach as dual topology. Ries et al suggested to use the term hybrid topology for (single coordinate) dual topology in the sense of Pearlman. We will attempt a disambiguation of the various nomenclatures below.

@@

Dual coordinate approaches with chemical topologies encompass a third group of methods, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category. In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other. In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand. Both copies are present simultaneously, but they do not occupy the same space. Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. However, DC approaches are also extremely prone to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% What simulation protocol should I choose   %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{\tocorange{What simulation protocol should I choose?}}
\section{\tocorange{Choice of simulation protocol and implementation of alchemical transformations}}
  \sbnote{Emilio suggested to make the title more general; this is my suggestion!}
\tocsectioncomment{Steering group: Possibly expand to include other methods group agrees on.}
\tocsectioncomment{@Stefan Boresch to restructure as per his suggestion to remove single/dual topology section and begin with a more fundamental discussion which might more naturally include all options e.g. ATM. @Stefan Boresch to mention best practices for dummy atoms? This might benefit from following @Emilio Gallicchio's changes to make the theory section more general.}
\tocsectioncomment{@Sheenam Khuttan / @Emilio Gallicchio to contribute section on ATM.}
\tocsectioncomment{@Vytas Gapsys / @Volunteer to add best practices for non-equilibrium calculations.}
\label{sec:simulation_protocol_choice}
Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (see Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative''\footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state and in that sense are relative, while neither computes an absolute free energy.} (see Fig.~\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
To use binding as a concrete example, in absolute binding free energy calculations, we compute the binding free energy of a ligand to an individual receptor relative to a standard reference concentration. In contrast, in relative binding free energy calculations, we compare the binding free energy of two related ligands to determine the potency difference.

@@

\paragraph{Topologies} In regular biomolecular simulations one studies physical systems or, in other words, all entities in the system, e.g., protein, ligand and waters, represent real molecules. For the following, it is helpful to classify such simulation systems as having a {\em chemical} topology. In relative free energy calculations, however, one has to transform, e.g., ligand A into ligand B etc. (see \sbnote{appropriate cross-reference!}). The required alchemical transformations can be realized in several ways, each with distinct advantages and disadvantages. In the first relative free energy calculations, these transformations were achieved with the help of chimeric entities, capable of representing both ligand A or ligand B. The important point here is that there is a single set of coordinates for this chimeric ligand, but its ``topology'' is flexible and can, for the same set of coordinates, represent either end state. We will consider this class of approaches as using an {\em alchemical} topology with a single set of coordinates. The construction of these chimeric intermediates quickly gets complicated and is difficult to automate; therefore, approaches have been developed in which both end states (e.g., ligand A and B) are present simultaneously. Thus, we have dual coordinates, but chemical topologies for each of the entities. In the following, we outline these approaches in more detail, together with additional considerations how the various approaches differ in practical use. While we will point out the respective advantages and disadavantages, we stress that the user's option in practice often are limited by the choice of simulation software.

@@

In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by first changing one of the hydrogens into a carbon. The site of this transformation will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'' (labelled Du in Fig.~\ref{fig:fig_topology}), which retain their bonded interactions but do not interact with the rest of the system.  Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final state; see below for further details. In a single topology calculation, atoms may change their type, ensuring minimal dummy atoms are created. This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}. At intermediate states, atoms behave as 'mixed' (M), either interpolating between two physical atoms, e.g., in the example, the hydrogen becomes a carbon, or between dummy and physical atoms.

\begin{figure}
    \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
    \caption{\textbf{Strategies for alchemical calculations: single coordinate single and dual topology vs.\ dual coordinate dual topology.} \textbf{Left}: A single coordinate single topology converts from one type of atom to another. Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms, using a soft-core interaction to improve overlap between the dummy atoms and the "real" atoms. At intermediate states, 'mixed' atoms (M) are present, either interpolating between physical atoms, e.g., a hydrogen to a carbon, or between physical and dummy atoms.  \textbf{Middle}: The single coordinate dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, typically using soft-core potentials for this. Here, 'mixed' intermediate atoms (M) always interpolate between physical and dummy atoms. Thus, 'mixed' atoms are used in both single coordinate approaches (single and dual topology), but the way the transformation occurs and the end states differ.
      %Following the arrow along the left and right illustrate the differences.
\textbf{Right}: In the dual coordinate dual topology approach both end states are present as separate molecules simultaneously.  Initially, benzyl alcohol is not interacting with the environment, at intermediate states the environment sees a mixture of the interactions of the two solutes, and, finally, benzene is not interacting with the environment. The two solutes are kept on top of each other by suitable restraints, indicated by the black ``spring'', and  the two molecules never see each other.  Figure adapted from the one used in V1 of this manuscript; adapted and extended by Sara Tkaczyk.
      % \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}
    }
    \label{fig:fig_topology}
\end{figure}

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works. Note that the two groups that are switched on or off never see each other. The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.

%\sbnote{This probably needs to be rewritten\ldots}
As far as we are aware, the distinction between (single coordinate) single and dual topology approaches has been made primarily for two main reasons. First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches. Additionally, historically, some simulation packages implemented only one approach and not the other, meaning that the distinction was functional. For an in-depth analyis of differences resulting from the use of a  single vs.\ a dual topology approach (single coordinate in both cases) see Ref.~\cite{Boresch_2002}.

@@

The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts (see below). %\sbnote{See below; where do we place the dummy atom discussion?}.
Some of these difficulties might be avoided by having the initial and final state present as separate molecules with separate coordinate sets. To the best of our knowledge, such an approach was first described by Axelsen and Li \cite{Axelsen_1998}. Both initial and final state are present simultaneously; the two copies never see each other and interactions with the environment are turned on/off as a function of $\lambda$. Harmonic restraints between corresponding atoms in the initial and final state copy prevent the two molecules (sets of atoms from drifting apart). %Unfortunately, Axelsen and Li referred to their ansatz also as dual-topology, though in practice it is quite different from Pearlman’s meaning of the term.
A recent implementation of this idea is Restraintmaker \cite{Ries_2022}, a tool to automatically assign the distance restraints needed to keep the two molecules / regions (one for each of the physical endstates) on top of each other. The idea is illustrated in the right branch of Fig.~\ref{fig:fig_topology}.  Somewhat unfortunately, Axelsen and Li, as well as Ries et al refer to their approach as dual topology. Ries et al suggested to use the term hybrid topology for (single coordinate) dual topology in the sense of Pearlman. We will attempt a disambiguation of the various nomenclatures below.

@@

Dual coordinate approaches with chemical topologies encompass a third group of methods, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category. In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other. %In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
In ATM, restraints are used to keep the bound and free ligands at some distance apart and in the same orientation, so that during the translation move the free ligand is placed into a position and orientation as similar as possible to that of the bound ligand. Both copies are present simultaneously, but they do not occupy the same space. Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} aoid this problem.
